Here we'll look at **Africa**:

* The least "stretched" by different projections,
* Interesting geometric properties like countries-inside-of-countries (South Africa, Gambia), and
* Will be helpful for comparing/contrasting different ways of defining "neighbors"/neighborhoods

---

## Getting the ISO3 Codes {.smaller}

:::: {.columns}
::: {.column width="50%"}

* Pulling my PhD card for just one slide...
* **"Country"** can be ambiguous/arbitrary (UK? England / Wales / Scotland / Northern Ireland?)
* **"Continent"** can be **even more** ambiguous/arbitrary
* $\implies$ **choose** a definition, **cite** it, use **ISO**, and be willing to **change** it on the basis of audience/peer review/global geopolitical changes/etc.
* Here we'll use the UNEP (UN Environment Program) definition of continents, from `rworldmap`

:::
::: {.column width="50%"}

```{r}
#| label: get-iso3
library(tidyverse)
library(rworldmap)
afr_countries <- rworldmap::countryRegions |>
  filter(GEO3major == "Africa") |>
  select(ISO3, ADMIN, GEO3major)
afr_countries |> head()
```

:::
::::


```{r}
#| label: get-geos
#| fig-align: center

#asia <- ne_countries(continent = "Asia")
usa <- ne_countries(country = "United States of America", scale = 10)
us_map <- mapview(usa, label = "geounit", legend = FALSE)
us_map
```

## Centroid

```{r}
#| label: us-centroid
library(sf)
usa_cent <- sf::st_centroid(usa)
us_map + mapview(usa_cent)
```

---

## Pairs of Geometries {.smaller}

```{r}
#| label: lesotho-sa
sa <- ne_countries(country = "South Africa", scale = 50)
lesotho <- ne_countries(country = "Lesotho", scale=50)
sf::st_relate(lesotho, sa)
sf::st_covers(lesotho, sa)
sf::st_covered_by(lesotho, sa)
mapview(sa) + mapview(lesotho)
```
