## HW1 Question 1 Generally

* "Simplest" / Most Efficient Representation

  > *"Indicate which of the seven geometries above would provide the **simplest** representation of the entity"*

* `GEOMETRYCOLLECTION` *could* represent any of the geographic entities in Q1, but would be overkill for representing e.g. a single point or line

## HW1 Question 1.8 Specifically {.smaller .leaflet-300 .crunch-title}

For the United Arab Emirates (UAE) data... we have what we call the **[Not-Paraguay-Problem]{style="font-variant: small-caps;"}**: Most countries, including the UAE, have a bunch of lil "pieces":

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: russia-pieces
ru_national_map <- ne_countries(type = "countries", country = "Russia", scale = "medium", returnclass = "sf")
mapview(ru_national_map)
```

:::
::: {.column width="50%"}

```{r}
#| label: uae-pieces
uae_national_map <- ne_countries(type = "countries", country = "United Arab Emirates", scale = "large", returnclass = "sf")
mapview(uae_national_map)
```

:::
::::

So, for Q1.8: Assume we're just trying to have the computer represent the **"mainland"** (the biggest contiguous landmass, with Dubai on it!)

---



## Public vs. Hidden Tests

* Public Tests are basically **"Quality Assurance Test"**
* Hidden Tests check for **correctness**